Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2022-06-07 inst/manuscript/manuscript_microxanox.Rmd
Fix crossrefs

# \textcolor{red}{TODO}
- **\textcolor{red}{update links}**
- **\textcolor{red}{update cross references}**
- **\textcolor{red}{check keywords}**
- **\textcolor{red}{add more references}**
- **\textcolor{red}{expand on Impact and Conclusion section}**
\pagebreak

@@

# Software description
> {>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<}

@@

## Software architecture
> {>>Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.<<}

@@

A typical simulation would look as shown in \@ref(fig:runsim_example).

@@

The function `plot_dynamics()` plots a single simulation run, as returned from the `run_simulation()` function. This function is only provided as a convenience function to provide a way to easily see the results of a simulation run. An example plot resulting from this function is shown in \@ref(fig:plot-dynamics).

```{r plot-dynamics, echo = FALSE,  include=TRUE, fig.cap='Results of a simulation run shown using the function $plot_dynamics()$. In this case, there were three strains per functional group, though strains within functional groups had identical properties in this example. CB_1 = cyanobacteria strain 1; SB_1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show the figure produce by $plot_dynamics()$, which produces figures not with all features ideal for publication.', fig.align="center", out.width = 350 }
knitr::include_graphics("figures/ug_three_strains_dynamics.png")
```

@@

> {>>This is the main section of the article and the reviewers weight the description here appropriately
Indicate in what way new research questions can be pursued as a result of the software (if any).
Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so).
Indicate in what way the software has changed the daily practice of its users (if so).
Indicate how widespread the use of the software is within and outside the intended user group.
Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<}

@@

An example of this functionality is given in the *Partial Reproduction* vignette, which we briefly show here (Figure \@ref(fig:uc1). The example is composed of a single simulation, at the beginning of which the system is in the oxic state with high abundance of cyanobacteria. Oyxgen diffusivity is then slowly decreased and eventually, around hour 30'000 the system switches to the anoxic state, with high abundance of both sulfur bacteria types. The oxygen diffusivity is then increased and at around hour 38'000 the system abruptly switches back to the oxic state.

@@

```{r uc1, echo = FALSE,  include=TRUE, fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity. In this simulation there is only one strain in each functional group. CB_1 = cyanobacteria strain 1; SB_1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show a figure adapted from the output of the $plot_dynamics()$  function.', fig.align="center", out.width = 350}
fig <- here::here("inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching.pdf")
#if (!file.exists(fig)){
  library(ggplot2)
  library(microxanox)
  library(dplyr)
  x <- readRDS(here::here("inst/manuscript/data/hlh_with_defaults.RDS"))
  p <- plot_dynamics(x)
  p[[1]] <- p[[1]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[2]] <- p[[2]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[3]] <- p[[3]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p_a <- p[[4]]
  p[[4]]$data <- p[[4]]$data %>% dplyr::filter(species != "a")
  p[[4]]$data$species <- p[[4]]$data$species %>%
    dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
  p[[4]] <- p[[4]] +
    xlab("time (h)") + 
    ylab(expression(atop(log[10](concentration),(µM)))) + 
    labs(colour = "substrate") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p_a$data <- p_a$data %>% dplyr::filter(species == "a")
  p_a$data$species <- p_a$data$species %>%
    dplyr::recode(a = "Oxygen diffusivity ")
  p_a <- p_a +
    ylab(expression(atop(Oxygen~diffusivity,(hr^{-1})))) + 
    labs(colour = "Environmental driver") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p <- p[[1]] / p[[2]] / p[[3]] / p[[4]] / p_a
  ggsave(
    filename = fig,
    plot = p,
    device = "pdf",
    width = 8,
    height = 8
  )  
#}
knitr::include_graphics(fig)
```

@@

The results show that the amount of hysteresis depends greatly on the combinations of organisms present (Figure \@ref(fig:uc2)). For example, with only the CB (cyanobacteria) present, there was no hysteresis. In contrast, the presence of both CB and SB (sulfate reducing bacteria) led to a large amount of hysteresis. (These results are also given in the *Partial Reproduction* vignette.)

@@

Figure \@ref(fig:uc3) shows a simulation with two species (strains) in each of the three functional groups. The ecosystem starts in the oxic state, though with relatively high abundance of each functional group. The strain of SB that is more tolerant to oxygen (SB_1) initially decreases in abundance, but then increases, and the other (SB_2) strain then becomes abundance and SB_1 declines. Furthermore, the cyanobacteria crash in abundance, and the system switches to the anoxic state. In contrast, if there are two identical strains with tolerance half way between those in Figure \@ref(fig:uc3) the ecosystem remains in the oxic state.

@@

# Conclusions
> {>>Set out the conclusion of this original software publication.<<}

>>>>

# \textcolor{red}{TODO}
- **\textcolor{red}{update links}**
- [x] ~~**\textcolor{red}{update cross references}**~~
- **\textcolor{red}{check keywords}**
- [x] ~~**\textcolor{red}{add more references}**~~
- **\textcolor{red}{expand on Impact and Conclusion section}**
\pagebreak

@@

# Software description
<!-- > {>>Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.<<} -->

@@

## Software architecture
<!-- > {>>Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.<<} -->

@@

A typical simulation would look as shown in \ref{fig:runsim_example}.

@@

The function `plot_dynamics()` plots a single simulation run, as returned from the `run_simulation()` function. This function is only provided as a convenience function to provide a way to easily see the results of a simulation run. An example plot resulting from this function is shown in \ref{fig:plot-dynamics}.

```{r plot-dynamics, echo = FALSE,  include=TRUE, fig.cap="Results of a simulation run shown using the function $plot_dynamics()$. In this case, there were three strains per functional group, though strains within functional groups had identical properties in this example. CB\textunderscore1 = cyanobacteria strain 1; SB\textunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show the figure produce by $plot_dynamics()$, which produces figures not with all features ideal for publication.", fig.align="center", out.width = 350 }
knitr::include_graphics("figures/ug_three_strains_dynamics.png")
```

@@

<!-- > {>>This is the main section of the article and the reviewers weight the description here appropriately -->
<!-- Indicate in what way new research questions can be pursued as a result of the software (if any). -->
<!-- Indicate in what way, and to what extent, the pursuit of existing research questions is improved (if so). -->
<!-- Indicate in what way the software has changed the daily practice of its users (if so). -->
<!-- Indicate how widespread the use of the software is within and outside the intended user group. -->
<!-- Indicate in what way the software is used in commercial settings and/or how it led to the creation of spin-off companies (if so).<<} -->

@@

An example of this functionality is given in the *Partial Reproduction* vignette, which we briefly show here (Figure \ref{fig:uc1}. The example is composed of a single simulation, at the beginning of which the system is in the oxic state with high abundance of cyanobacteria. Oyxgen diffusivity is then slowly decreased and eventually, around hour 30'000 the system switches to the anoxic state, with high abundance of both sulfur bacteria types. The oxygen diffusivity is then increased and at around hour 38'000 the system abruptly switches back to the oxic state.

@@

```{r uc1, echo = FALSE,  include=TRUE, fig.cap='The temporal dynamics of the ecosystem model when an environmental condition (here parameter *a*, the oxygen diffusivity) changes. Plot of the stable states of the simulation runs under different oxygen diffusivity. In this simulation there is only one strain in each functional group. CB\textunderscore1 = cyanobacteria strain 1; SB\textunderscore1 = sulfur reducing bacteria stain 1; PB = phototrophic sulfur bacteria strain 1. Here we show a figure adapted from the output of the $plot_dynamics()$  function.', fig.align="center", out.width = 350}
fig <- here::here("inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching.pdf")
#if (!file.exists(fig)){
  library(ggplot2)
  library(microxanox)
  library(dplyr)
  x <- readRDS(here::here("inst/manuscript/data/hlh_with_defaults.RDS"))
  p <- plot_dynamics(x)
  p[[1]] <- p[[1]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[2]] <- p[[2]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p[[3]] <- p[[3]] + 
    xlab(NULL) + 
    ylab(expression(atop(log[10](density),(cells~L^{-1})))) +
    theme_bw()
  p_a <- p[[4]]
  p[[4]]$data <- p[[4]]$data %>% dplyr::filter(species != "a")
  p[[4]]$data$species <- p[[4]]$data$species %>%
    dplyr::recode(SO = "Oxidised sulfur", O = "Oxygen", SR = "Sulfide", P = "Phosphorus")
  p[[4]] <- p[[4]] +
    xlab("time (h)") + 
    ylab(expression(atop(log[10](concentration),(µM)))) + 
    labs(colour = "substrate") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p_a$data <- p_a$data %>% dplyr::filter(species == "a")
  p_a$data$species <- p_a$data$species %>%
    dplyr::recode(a = "Oxygen diffusivity ")
  p_a <- p_a +
    ylab(expression(atop(Oxygen~diffusivity,(hr^{-1})))) + 
    labs(colour = "Environmental driver") +
    theme_bw() +
    xlab(expression(time~(hours)))
  p <- p[[1]] / p[[2]] / p[[3]] / p[[4]] / p_a
  ggsave(
    filename = fig,
    plot = p,
    device = "pdf",
    width = 8,
    height = 8
  )  
#}
knitr::include_graphics(fig)
```

@@

The results show that the amount of hysteresis depends greatly on the combinations of organisms present (Figure \ref(figL:uc2)). For example, with only the CB (cyanobacteria) present, there was no hysteresis. In contrast, the presence of both CB and SB (sulfate reducing bacteria) led to a large amount of hysteresis. (These results are also given in the *Partial Reproduction* vignette.)

@@

Figure \ref{fig:uc3} shows a simulation with two species (strains) in each of the three functional groups. The ecosystem starts in the oxic state, though with relatively high abundance of each functional group. The strain of SB that is more tolerant to oxygen (SB_1) initially decreases in abundance, but then increases, and the other (SB_2) strain then becomes abundance and SB_1 declines. Furthermore, the cyanobacteria crash in abundance, and the system switches to the anoxic state. In contrast, if there are two identical strains with tolerance half way between those in Figure \ref{fig:uc3} the ecosystem remains in the oxic state.

@@

# Conclusions
<!-- > {>>Set out the conclusion of this original software publication.<<} -->
